### 2017-01-06 paper/chapters/sigmoid_belief_nets.tex
more updates

\section{Sigmoid Belief Networks}
\label{sec:sbn}
\begin{figure}[htb]
\centering
\input{media/sbn_architecture}
  \caption{Conceptual architecture of belief nets}
  \label{fig:sbn_arch}
  \medskip
  \small
  Bayesian networks, also called belief nets can be described as an acyclic directed graph where blue nodes indicate hidden variables while red nodes represent observed variables.
  Arrows show stochastic dependency. The dashed arrow indicates that it is possible in the original definition, even though most commonly the hidden variables are divided into layers.
\end{figure}

Sigmoid Belief Networks (SBN) have been proposed in 1992 by R.M. Neal \cite{neal:1992} representing a specific type of bayesian networks \cite{pearl:1985}.
Figure \ref{fig:sbn_arch} shows a generic belief network as a graphical model with each state influenced by its ancestors.
In the case of SBNs however, all states are binary and the activation function is the sigmoid function. They are one of the earliest neural networks used for generative modelling, predating all other presented models in this article.
%As described in the overview above, SBNs contain a number of binary states $s$, most commonly divided into many layers.
Each state $s_i$ is a random variable in the bayesian framework which means it can be observed, hidden or unknown representing different states of knowledge. As such it can also be interpreted with a probability distribution.
\begin{equation}
  \label{eq:sbn_node}
p(s_i) = \sigma\bigg(\sum_{j<i}W_{i,j}s_j+b_i\bigg)
\end{equation}
In equation \ref{eq:sbn_node}, $\sigma$ denotes the sigmoid function $\frac{1}{1 + e^{-x}}$ and $W_{i,j}$ the connection weights from state $s_j$ to $s_i$.

To draw samples from this network, it is possible to simply evaluate equation \ref{eq:sbn_node} for each visible node.
However prior to that, the SBN first needs to learn how to generate good samples which includes the \emph{learning} and the \emph{inference} problem.\\
Learning means in this context to adjust the weights and biases of the network to be more likely to generate previously observed input.
Meanwhile statistical inference is tasked to infer the states of the hidden nodes based on the input.\\

Due to each state being statistically depend on all ancestors, inference means to evaluate all possible configurations of every previous node.
Especially in deep networks, this kind of computation is infeasible as it requires evaluating exponentially many configurations.\\

>>>>

\section{Sigmoid Belief Networks}
\label{sec:sbn}
\begin{figure}[htb]
\centering
\input{media/sbn_architecture}
  \caption{Conceptual architecture of belief nets}
  \label{fig:sbn_arch}
  \medskip
  \small
  Bayesian networks, also called belief nets are an acyclic directed graph where blue nodes indicate hidden variables and red nodes represent observed variables.
  Arrows indicate stochastic dependency.
  Even though most belief nets are divided into many layers in the context of deep learning, the original definition does not specify that and thus indicates the dashed arrow the possibility of inter-layer dependencies.
\end{figure}

Sigmoid Belief Networks (SBN) were proposed in 1992 by R.M. Neal \cite{neal:1992} representing a specific type of bayesian networks \cite{pearl:1985} shown in figure \ref{fig:sbn_arch}
In the case of SBNs however, all states are binary and the activation function is the sigmoid function. They are one of the earliest neural networks used for generative modeling, predating all other presented models in this article.
%As described in the overview above, SBNs contain a number of binary states $s$, most commonly divided into many layers.
Each state $s_i$ is a random variable in the bayesian framework which means it can be observed, hidden or unknown representing different states of knowledge. As such it can also be interpreted with a probability distribution.
\begin{equation}
  \label{eq:sbn_node}
p(s_i) = \sigma\bigg(\sum_{j<i}W_{i,j}s_j+b_i\bigg)
\end{equation}
In equation \ref{eq:sbn_node}, $\sigma$ denotes the sigmoid function $\frac{1}{1 + e^{-x}}$ and $W_{i,j}$ the connection weights from state $s_j$ to $s_i$.

To draw samples from this network, it is possible to simply evaluate equation \ref{eq:sbn_node} for each visible node.
However prior to that, the SBN first needs to learn how to generate good samples which includes the \emph{learning} and the \emph{inference} problem.\\
Learning means in this context to adjust the weights and biases of the network to be more likely to generate previously observed input.
Meanwhile statistical inference is tasked to infer the states of the hidden nodes based on the states in the visible layer.\\

Due to each state being statistically depend on all ancestors, inference means to evaluate all possible configurations of every previous node.
Especially in deep networks, this kind of computation is infeasible as it requires computing exponentially many configurations.\\

### 2017-01-06 paper/chapters/vae.tex
more updates

\section{Variational Autoencoders}
\label{sec:vae}
Variational autoencoders (VAE) have been developed concurrently in 2014 by Kingma et al. \cite{vae:2014} and Rezende et al. \cite{dlgm:2014}.
%The proposed framework resembles the encoder-decoder architecture, using a probabilistic encoder and decoder.
The general idea of VAEs is to have an encoder-decoder architecture where inputs $x$ gets mapped to latent variables $z$ and from there to the reconstruction $~x$.

@@

However, in addition to this autoencoder structure an additional regularizer forces the latent space to have a specific simple form.
To achieve this, variational-based methods are used to output a distribution on the latent space instead of directly emitting samples.

The probabilistic encoder generally implemented as feed-forward network receives input from the dataset and outputs parameters of a distribution. For example the parameters for a gaussian distribution with diagonal covariance, in which case the networks emits $(\mu,\Sigma)$.\\

Using gradient-based optimization techniques enables the network to be trained to be similar to the true posterior $p(z|x)$.
In addition to this inference network there is a generator network which transforms points $z$ in the latent space back to the data space while trying to minimize the reconstruction error.

@@

Formally, the described probabilitic encoder emits variational parameters $\phi$ used to define the probability distribution $q_\phi(z|x)$.
Equally for the probabilistic decoder which constructs a distribution $p_\theta(x|z)$ with the parameters $\theta$.\\

@@

The key insight on how the variational autoencoder is able to optimize $q_\phi(z|x)$ to approach the true data posterior distribution $p(z|x)$ is in defining a variational lower bound which is used to push the approximated $q_\phi(z|x)$ towards $p(z|x)$. We will discuss this lower bound $\mathcal{L}$ in the following section in detail.

@@

These intractable posteriors unfortunately arise in almost all interesting models and there are mostly two different approaches to approximate the posterior distribution.
% MCMC
Using Markov chain Monte Carlo (MCMC) methods, which construct a Markov chain with the desired attributes. The advantage of this approach is that is guaranteed to asympotically approach the desired posterior, however it is difficult to reason about the convergence rate. For deep neural networks, running MCMC in the inner loop of optimization techniques is impractical.\\
% variational inference
The other approach which is using variational inference which picks a specific distribution restricted to a family of distributions matching the posterior most closely.
Advantages using this approach rather than MCMC is the greater speed at which the computation is performed.
However it is not guaranteed to even approximate the posterior accurately, if the true posterior can not be expressed as a distribution the family.

@@

\subsubsection{Kullback-Leibler Divergence}
In order to approximate one distribution $Q$ to another distribution $P$, a measurement of difference.
One commonly used measurement is called Kullback-Leibler divergence (short KL-divergence) which enables the variational inference methods to choose appropiate distributions.
%\paragraph{The Kullback-Leibler Divergence} (KL divergence) is a measurement for the difference between two probability distributions. $\mathcal{D}_{\mathrm{KL}}(P \| Q)$ can informally described as the amount of information which is lost when using $Q$ to represent $P$.
For discrete probability distributions $P$ and $Q$, the KL divergence from $Q$ to $P$ is defined\cite{kl_div:1951} as:
\begin{equation}
  \mathcal{D}_{\mathrm{KL}}(P || Q) = \sum_{i} P(i) \log \frac{P(i)}{Q(i)}
\end{equation}

@@

\subsection{Objective Function}
%The objective function of the VAE consists of the reconstruction error and the regularizer.
%$\mathcal{L}$ learns both the encoder as well as the decoder with their respective parameters $\phi$ and $\theta$.
% probabilistic encoder $q_\phi(z|x)$ (produces z values from which x could've been generated)
% probabilistic decoder $p_\theta(x|z)$
%We will discuss both terms in detail below following the derivation of the lower variational bound $\mathcal{L}$ and the rewritten objective function.
Following the thoughts earlier, we wish to derivate an objective function which can be used to perform gradient-based optimization in the model.\\
We construct the approximate posterior $q_\phi(z|x)$ which we want to approach the true data posterior $p(z|x)$.
When using the KL divergence as measurement for difference between two distributions the goal for which we want to optimize becomes the following.
\begin{equation}
  q_\phi^{*}(z|x) = \arg\min_{\phi} \mathcal{D}_{\mathrm{KL}} \big(q_\phi(z|x) || p(z|x)\big)
\end{equation}

However we are not able to optimize for that goal directly, as the intractable evidence $p(x)$ appears during the expansion of the posterior $p(z|x)$ to $\frac{p(x|z)p(z)}{p(x)}$.\\
The denominator $p(x)$ can be computed by marginalizing out the hidden variables $z$: 
\begin{equation}
  \label{eq:integral}
p(x) = \int p(x,z) p(z) dz
\end{equation}

Computing the integral in equation \ref{eq:integral} requires the evaluation of every possible configuration over the hidden variables $z$ and therefore intractable.\\\\

Instead of directly optimizing for the true posterior, we construct an evidence lower bound (ELBO) $\mathcal{L}$ which we can maximize which is equivalent to minimizing the KL divergence between $q_\phi(z|x)$ and $p(z|x)$.\\

@@

With the example using handwritten digits it appears why this regularizer is important for the model to learn generalized representations in latent space.
Suppose we have two data points from the dataset which both represent the same handwritten digit.
Ideally, our model would learn the representations in latent space to be fairly close to ensure the model has learned to generalize the semantics. Without $\mathcal{L}_{reg}$ the model would assign each data example a different representation in latent space with a distribution collapsing to that single point. This behaviour is undesired and is as such penalitized.

@@

\paragraph{Reconstruction error $\mathcal{L}_{rec}$} measures how well the decoder reconstructs the input data using the latent space $z$.
This error is measures using the negative log-likelihood of the conditional probability distribution $p_\theta(x_i|z)$ where $z$ is sampled from $q_\phi(z | x_i)$.\\
$\mathcal{L}_{\mathrm{rec}}$ is necessary to force the encoder to produce latent variables which can be used to reconstruct the input data well.

@@

    \subcaptionbox{with reparameterization applied}{\includestandalone[height=0.4\linewidth]{media/reparameterization_trick_with}}\\
  \end{tabular}
  \caption{reparameterization trick}
  \medskip
  \small
  Function $f_\theta(x)$ with inputs $x$ and (variational) parameters $\theta$ shown as directed graphical model.\\
  Left without and on the right with the reparameterization trick applied.\\
  Solid arrows show the forward pass and dashed arrows the error backpropagation.
  Rectangular nodes represent deterministic functions while circles indicate stochastic variables.
  %Note that backpropagation of gradients through stochastic nodes is problematic and adds bias (verify!)
\end{figure}

% more math:
% http://stats.stackexchange.com/questions/199605/how-does-the-reparameterization-trick-for-vaes-work-and-why-is-it-important
To be able to backpropagate the loss function through the VAE, it has to be differentiable and deterministic.
Because we add noise to the encoding, the gradients can't be computed directly. In order to circumvent this restriction, the so-called "reparameterization trick" is applied.
Instead of drawing $z ~ \mathcal{N}(\mu(x), \Sigma(x))$, we sample an auxiliary variable $\epsilon$ from $\mathcal{N}(0, I)$ which we then transform with equation \ref{eq:rep_trick} into $z$.
\begin{equation}
  \label{eq:rep_trick}
  z = \mu(x) + \Sigma^{1/2}(x)*\epsilon
\end{equation}
This allows us to compute the gradient of the loss function and backpropagate through the entire model and only have the stochastic variable $\epsilon$ as input.\\\\

The core idea of VAEs is to match a distribution $q(x)$ to the desired $p(x)$ by enforcing a lower variational bound on $q(x)$ in a way that it seeks to match $p(x)$.
In contrast to other autoencoders, VAEs behave differently due to this lower bound but they resemble the architecture of traditional autoencoders (sparse, denoising).

We use VAEs when we have a complicated distribution $p_\theta(x,z)$ with unknown latent variables $z$.
The prior distribution $p(z)$ over the latent structure is a centered isotropic(?) multivariate Gaussian denoted by $\mathcal{N}(z;0, I)$.
The constructed architecture uses an probabilistic encoder $q_\theta(z|x)$ and a probabilistic decoder $p_\theta(x|z)$ in a form of neural networks, in the original paper MLP are used\cite{vae:2014} but there are various extensions (see \ref{sub:vae_extensions}, TODO).
Because the posterior is intractable ($p_\theta(x)$), we instead approximately maximize the lower variational bound $L(\theta,\phi;x)$.\\
\begin{equation}
  \mathcal{L}(\theta,\phi;x) = -D_{KL}(q_\theta(z|x)||p(z)) + \mathbb{E}_{q_\theta(z|x)}[log_{p_\theta}(x|z)]
\end{equation}

@@

\subsection{Learning}
\label{sub:vae_learning}
The probabilistic encoder $q_\phi(z|x)$ and decoder $p_\theta(x|z)$ are commonly implemented using a deep neural network.
To allows analytical computation of the ELBO, the prior on the latent space $z$ is often choosen to be the isotropic gaussian $p(z) = \mathcal{N}(0,1)$.

@@

\subsubsection{Conditional VAE \cite{cvae:2014} \cite{cvae_2:2015}}
\label{ssub:vae_conditional_vae}
\input{chapters/vae/cvae}

@@

\subsubsection{Deep Latent Gaussian Model}
%\begin{figure*}[t!]
    %\centering
    %\includegraphics[height=7cm]{media/dlgm_computational_graph-crop.pdf}
    %\caption{VAE with multiple stochastic layers as a deep latent gaussian model \cite{dlgm:2014}}
    %\label{fig:dlgm_model}
    %\medskip
    %\small
    %This figure shows the computational graph in a deep latent gaussian model. Red arrows indicate the forward pass and black arrows the backward pass. Meanwhile represent j
%\end{figure*}
Up until now the presented variational autoencoder only has one stochastic latent layer. However in the original paper by Rezende et al. \cite{dlgm:2014} the authors have shown a variational autoencoder using multiple stochastic layers with gaussian priors in each layer.
Note that the deterministic non-linear transformations between each layer still can be represented using deep neural networks.
This enables the model to learn richer, more hierarchical structures captured in each latent layer.

@@

\subsubsection{Importance Weighted Autoencoder \cite{iwae:2015}}
\label{ssub:vae_importance_weighted_autoencoder}
\input{chapters/vae/iwae}

\subsubsection{Deep Recurrent Attention Writer \cite{draw:2015}}
\label{ssub:vae_deep_recurrent_attention_writer}
\input{chapters/vae/draw}

>>>>

\section{Variational Autoencoders}
\label{sec:vae}
Variational autoencoders (VAE) have been developed concurrently in 2014 by Kingma et al.\cite{vae:2014} and Rezende et al.\cite{dlgm:2014}.
%The proposed framework resembles the encoder-decoder architecture, using a probabilistic encoder and decoder.
The general idea of VAEs is to have an encoder-decoder architecture where input $x$ gets mapped to a specific distribution defined in the latent space $z$ from which we sample in order to reconstruct the input.

@@

%However, in addition to this autoencoder structure an additional regularizer forces the latent space to have a specific simple form.
%To achieve this, variational-based methods are used to output a distribution on the latent space instead of directly emitting samples.

The probabilistic encoder generally implemented as feed-forward network receives input from the dataset and outputs parameters of a distribution. For example the parameters for a gaussian distribution with diagonal covariance, in which case the networks emits the mean $\mu$ and $\sigma$ used to define the gaussian distribution $\mathcal{N}(\mu,\sigma \times I)$, where $I$ is the identity matrix.\\

Using gradient-based optimization techniques enables the encoder network to be trained to output distributions similar to the true posterior $p(z|x)$.
In addition to this inference network there is a generator network which transforms points $z$ in the latent space back to the data space while trying to minimize the reconstruction error.

@@

Properly defined emits the described probabilitic encoder variational parameters $\phi$ used to define the probability distribution $q_\phi(z|x)$.
Equally for the probabilistic decoder which constructs a distribution $p_\theta(x|z)$ with the parameters $\theta$.
This architecture is shown in figure \ref{fig:vae_architecture} with a one-dimensonal gaussian prior.
\\

@@

The key insight on how the variational autoencoder is able to optimize $q_\phi(z|x)$ to approach the true data posterior distribution $p(z|x)$ is in defining a variational lower bound which is used to push the approximated $q_\phi(z|x)$ towards $p(z|x)$. We will discuss this lower bound $\mathcal{L}$ and its derivation in detail in the following sections.

@@

These intractable posteriors unfortunately arise in almost all interesting models and there are mostly two different approaches to approximate the posterior distribution.
% MCMC
Using Markov chain Monte Carlo (MCMC) methods, which construct a Markov chain with the desired attributes matching those of the posterior. The advantage of this approach is that is guaranteed to asympotically approach the desired posterior, however it is difficult to reason about the convergence rate. For deep neural networks, running MCMC in the inner loop of optimization techniques is impractical.\\
% variational inference
Using variational inference is the second approach, used in the variational autoencoder, which picks a specific distribution restricted to a family of distributions matching the posterior most closely.
Advantages using this approach rather than MCMC is the greater speed at which the computation is performed.
However it is not guaranteed to approximate the posterior accurately, if the true posterior can not be approximately expressed as a distribution of the family.

@@

\subsubsection{Kullback-Leibler Divergence}
In order to approximate one distribution $Q$ to another distribution $P$, there needs to be a measurement how close $Q$ is to $P$.
One commonly used measurement is called Kullback-Leibler divergence (short KL divergence) which enables the variational inference methods to choose appropiate distributions.
%\paragraph{The Kullback-Leibler Divergence} (KL divergence) is a measurement for the difference between two probability distributions. $\mathcal{D}_{\mathrm{KL}}(P \| Q)$ can informally described as the amount of information which is lost when using $Q$ to represent $P$.
For discrete probability distributions $P$ and $Q$, the KL divergence from $Q$ to $P$ is defined\cite{kl_div:1951} in equation \ref{eq:kl_div_disc}:
\begin{equation}
  \label{eq:kl_div_disc}
  \mathcal{D}_{\mathrm{KL}}(P || Q) = \sum_{i} P(i) \log \frac{P(i)}{Q(i)}
\end{equation}
For continuous distributions $P$ and $Q$ the KL divergence is defined similarly in equation \ref{eq:kl_div_cont}.
\begin{equation}
  \label{eq:kl_div_cont}
  \mathcal{D}_{\mathrm{KL}}(P || Q) = \int P(i) \log \frac{P(i)}{Q(i)} di
\end{equation}

@@

\subsection{Objective Function}
%The objective function of the VAE consists of the reconstruction error and the regularizer.
%$\mathcal{L}$ learns both the encoder as well as the decoder with their respective parameters $\phi$ and $\theta$.
% probabilistic encoder $q_\phi(z|x)$ (produces z values from which x could've been generated)
% probabilistic decoder $p_\theta(x|z)$
%We will discuss both terms in detail below following the derivation of the lower variational bound $\mathcal{L}$ and the rewritten objective function.
Following the thoughts earlier, we wish to derivate an objective function which can be used to perform gradient-based optimization in the model.\\
We construct the approximate posterior $q_\phi(z|x)$ which we want to approach the true data posterior $p(z|x)$.
When using the KL divergence as measurement for difference between two distributions the goal for which we want to optimize becomes the following.
\begin{equation}
  \label{eq:vae_obj}
  q_\phi^{*}(z|x) = \arg\min_{\phi} \mathcal{D}_{\mathrm{KL}} \big(q_\phi(z|x) || p(z|x)\big)
\end{equation}

However we are not able to optimize equation \ref{eq:vae_obj} directly, as the intractable evidence $p(x)$ appears during the expansion of the posterior $p(z|x)$ to $\frac{p(x|z)p(z)}{p(x)}$.\\
The denominator $p(x)$ can be computed by marginalizing out the hidden variables $z$: 
\begin{equation}
  \label{eq:integral}
p(x) = \int p(x,z) p(z) dz
\end{equation}

Computing the integral in equation \ref{eq:integral} requires the evaluation of every possible configuration over the hidden variables $z$ and is therefore intractable.\\\\

Instead of directly optimizing for the true posterior, we construct an evidence lower bound (ELBO) $\mathcal{L}$ which we can maximize which is equivalent to minimizing the KL divergence between $q_\phi(z|x)$ and $p(z|x)$ without explicity evaluating the evidence.\\

@@

Using the example of handwritten digits it appears why this regularizer term $\mathcal{L}_{reg}$ is important for the model to learn generalized representations in latent space.
Suppose we have two data points from the dataset which both represent the same handwritten digit.
Ideally, our model would learn representations in latent space to be fairly close to ensure the model has learned to generalize the semantics. Without $\mathcal{L}_{reg}$ the model would assign each data example a different representation in latent space with a distribution collapsing to that single point. This behaviour is undesired and is as such penalitized.

@@

\paragraph{Reconstruction error $\mathcal{L}_{rec}$} measures how well the decoder reconstructs the input data using the latent space $z$.
This error is measured using the negative log-likelihood of the conditional probability distribution $p_\theta(x|z)$ where $z$ is sampled from $q_\phi(z | x)$.\\
$\mathcal{L}_{\mathrm{rec}}$ is necessary to force the encoder to produce latent variables which can be used to reconstruct the input data well.

@@

    \subcaptionbox{with reparameterization applied}{\includestandalone[height=0.4\linewidth]{media/reparameterization_trick_with}}\\
  \end{tabular}
  \caption{reparameterization trick}
  \label{fig:rep_trick}
  \medskip
  \small
  Function $f_\theta(x)$ with inputs $x$ and (variational) parameters $\theta$ as computational graph.
  Left without and on the right with the reparameterization trick applied.\\
  Solid arrows show the forward pass and dashed arrows the error backpropagation using gradients.
  Rectangular nodes represent deterministic functions while circles indicate stochastic variables.
  %Note that backpropagation of gradients through stochastic nodes is problematic and adds bias (verify!)
\end{figure}

% more math:
% http://stats.stackexchange.com/questions/199605/how-does-the-reparameterization-trick-for-vaes-work-and-why-is-it-important
To be able to backpropagate the loss function through the VAE, it has to be differentiable and deterministic.
However, because we only sample from the constructed distribution $q_\phi(z|x)$, we cannot backpropagate through the node due to the stochastic nature.
In order to circumvent this restriction, the so-called "reparameterization trick" is applied.
Instead of drawing $z$ directly from $q_\phi(z|x)$, we first sample first an auxiliary variable $\epsilon$ from the isotropic gaussian $\mathcal{N}(0, I)$. Then we apply equation \ref{eq:rep_trick} to transform $\epsilon$ into $z$.
\begin{equation}
  \label{eq:rep_trick}
  z = \mu(x) + \Sigma^{1/2}(x)*\epsilon
\end{equation}
Note that this transformation is entirely deterministic and as such allows for backpropagation through the transformation into the variational parameters $\phi$.

Figure \ref{fig:rep_trick} expresses the reparameterization trick applied to an arbitrary function $f$ with the parameters $\theta$.\\
Using this trick, the whole VAE model can be trained using gradient-based optimization techniques.\\\\

Note further that in most common architectures, the distribution class enforced on the latent space is gaussian, there exists similar reparameterization techniques allowing different probability distributions on $z$.
%The core idea of VAEs is to match a distribution $q(x)$ to the desired $p(x)$ by enforcing a lower variational bound on $q(x)$ in a way that it seeks to match $p(x)$.
%In contrast to other autoencoders, VAEs behave differently due to this lower bound but they resemble the architecture of traditional autoencoders (sparse, denoising).

%We use VAEs when we have a complicated distribution $p_\theta(x,z)$ with unknown latent variables $z$.
%The prior distribution $p(z)$ over the latent structure is a centered isotropic(?) multivariate Gaussian denoted by $\mathcal{N}(z;0, I)$.
%The constructed architecture uses an probabilistic encoder $q_\theta(z|x)$ and a probabilistic decoder $p_\theta(x|z)$ in a form of neural networks, in the original paper MLP are used\cite{vae:2014} but there are various extensions (see \ref{sub:vae_extensions}, TODO).
%Because the posterior is intractable ($p_\theta(x)$), we instead approximately maximize the lower variational bound $L(\theta,\phi;x)$.\\
%\begin{equation}
  %\mathcal{L}(\theta,\phi;x) = -D_{KL}(q_\theta(z|x)||p(z)) + \mathbb{E}_{q_\theta(z|x)}[log_{p_\theta}(x|z)]
%\end{equation}

@@

\subsection{Learning}
\label{sub:vae_learning}
The probabilistic encoder $q_\phi(z|x)$ and decoder $p_\theta(x|z)$ are commonly implemented using deep neural networks.
To allow analytical computation of the ELBO, the prior on the latent space $z$ is often choosen to be the isotropic gaussian $p(z) = \mathcal{N}(0,1)$.

@@

\subsubsection{Conditional VAE \cite{cvae:2014} \cite{cvae_2:2015}}
\label{ssub:cvae}
\input{chapters/vae/cvae}

@@

\subsubsection{Deep Latent Gaussian Model}
%\begin{figure*}[t!]
    %\centering
    %\includegraphics[height=7cm]{media/dlgm_computational_graph-crop.pdf}
    %\caption{VAE with multiple stochastic layers as a deep latent gaussian model \cite{dlgm:2014}}
    %\label{fig:dlgm_model}
    %\medskip
    %\small
    %This figure shows the computational graph in a deep latent gaussian model. Red arrows indicate the forward pass and black arrows the backward pass. Meanwhile represent j
%\end{figure*}
Up until now the presented variational autoencoder framework has only considered one stochastic latent layer. However in the original paper by Rezende et al. \cite{dlgm:2014} the authors have shown a variational autoencoder using multiple stochastic layers with gaussian priors in each layer.
They called this model the deep latent gaussian model (DLGM) which can be understood as a generalization of the variational autoencoder.
Note that the deterministic non-linear transformations between each layer still can be represented using deep neural networks.\\
Using multiple stochastic layers enables the model to learn more hierarchical structured dependencies from the dataset.

@@

\subsubsection{Importance Weighted Autoencoder}
\label{ssub:vae_importance_weighted_autoencoder}
\input{chapters/vae/iwae}

\subsubsection{Deep Recurrent Attention Writer}
\label{ssub:vae_deep_recurrent_attention_writer}
\input{chapters/vae/draw}

### 2017-01-06 paper/chapters/vae/cvae.tex
more updates

Instead of using VAEs in a completely unsupervised manner, oftentimes learning from a dataset where a small subset of datapoints has corresponding labels.
Techniques learning from such datasets are called semi-supervised learning and can improve performance significantly.\\

@@

For datapoints $x$ where labels $y$ are available.
\begin{align}
  \label{eq:cvae_labels}
  \log p_\theta(x) &\geq \mathbb{E}_{q_\phi(z|x,y)} \bigg[log p_\theta(x|y,z) + \log p_\theta(y) + \log p(z) - \log q_\phi(z|x,y)\bigg]
\end{align}

>>>>

Instead of using VAEs in a completely unsupervised manner on a unlabeled dataset, it is also possible to learn from datasets where a small subset of datapoints has corresponding labels while most datapoints are still unlabeled.
Techniques learning from such datasets are called semi-supervised learning and can improve performance significantly.\\

@@

For datapoints $x$ where labels $y$ are available.
\begin{align}
  \label{eq:cvae_labels}
  \log p_\theta(x) &\geq \mathbb{E}_{q_\phi(z|x,y)} \bigg[\log p_\theta(x|y,z) + \log p_\theta(y) + \log p(z) - \log q_\phi(z|x,y)\bigg]
\end{align}

### 2017-01-06 paper/chapters/vae/iwae.tex
zZzZzzz almost done
more updates

% write about multiple stochastic layers
Until now, the presented model includes only one stochastic layer.
\cite{dlgm:2014} have presented a model for learning through multiple stochastic layers which can be easily applied to the VAE framework as shown in \ref{fig:vae_multiple_layers}.

@@

% write about IWAE, extends these k passes with importance weights
Importance Weighted Autoencoders (IWAE) extends on the VAE framework using $k$-sample importance weighting of the log-likelihood.
% add intuition for IWAE, ruslan explained it very well during his DL summer school talk! (talk_Montreal_2016_Salakhutdinov.pdf)
Using importance weighting can be informally interpreted as putting more weight on data with high likelihood.

The authors of the papers argue that using more samples will only tighten the lower variational bound.
$$
\log p(x) \geq \mathcal{L}_{k+1}(x) \geq \mathcal{L}_{k}(x)
$$

Additionally, if $\frac{p(h,x)}{q(h|x)}$ is bounded, then the model is guaranteed to reach the optimimum:
$$
\lim_{k \rightarrow \infty} \mathcal{L}_{k}(x) = \log p(x)
$$

Of course, using more samples will result in higher computational cost.

\vspace{5cm}

>>>>

% write about multiple stochastic layers

@@

% write about IWAE, extends these k passes with importance weights
Importance Weighted Autoencoders (IWAE\cite{iwae:2015}) extends on the VAE framework using $k$-sampled importance weighting of the log-likelihood.
% add intuition for IWAE, ruslan explained it very well during his DL summer school talk! (talk_Montreal_2016_Salakhutdinov.pdf)
Using importance weighting can be informally interpreted as putting more weight on data with high likelihood.

Improving on the theoretical results in the VAE framework, using IWAE with more samples will tighten the lower variational bound and therefore improve the overall performance of the model.
\begin{equation*}
\log p(x) \geq \mathcal{L}_{k+1}(x) \geq \mathcal{L}_{k}(x)
\end{equation*}

Additionally, if $\frac{p(h,x)}{q(h|x)}$ is bounded, then the model is guaranteed to reach the optimimum as $k$ approaches infinity in equation \ref{eq:iwae}
\begin{equation}
\label{eq:iwae}
\lim_{k \rightarrow \infty} \mathcal{L}_{k}(x) = \log p(x)
\end{equation}

However, using more samples will result in higher computational cost and thus should be carefully evaluated.

%\vspace{5cm}
